---
title: "Code comparison with Seurat"
author: "Stefano Mangiola"
date: "`r Sys.Date()`"
package: tidyseurat
output:
  html_vignette:
    toc_float: true
vignette: >
  %\VignetteEngine{knitr::knitr}
  %\VignetteIndexEntry{Code comparison with Seurat}
  %\usepackage[UTF-8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Filtering

Just as a quick example, we show how to filter cells based on sample size (number of cells) and UMAP location. You can imagine a similar example with more meaningful criteria.

tidyseurat

```{r}
library(dplyr)
library(Seurat)
library(tidyseurat)
data("pbmc_small")
seurat_object = pbmc_small

```


```
{r}
seurat_object %>%
  add_count(file) %>%
  filter(PC_1 > 0 & n > 40)
```

Seurat

```
{r}
pca_emb = Embeddings(object = seurat_object, reduction = "pca")
cell_pca = rownames(pca_emb[pca_emb[,1]>0,])
n =
  seurat_object[[]] %>%
  add_count(file) %>%
  pull(n)
seurat_object = AddMetaData( object = seurat_object, metadata = n, col.name = 'n')
seurat_object %>%
  subset(cells = cell_pca) %>%
  subset( n < 40)

```


# Case study code comparison


## Calculate gamma-delta signature and plot

tidyseurat

## Gate cells based on expression

tidyseurat

## Balanced sampling and UMAP plot with cell cyle and features

tidyseurat

```
{r}

seurat_object %>%
  
  # Balanced sampling
  add_count(sample) %>%
  mutate(min_n = min(n)) %>%
  nest(data = -c(sample, n)) %>%
  mutate(data = map2(data, min_n, ~ .x %>% sample_n(.y))) %>%
  unnest(data) %>%
  
  # Extract features
  join_features(c("CD3G", "LYZ", "C2", "HLA-A"), shape="wide") %>%
  pivot_longer(c(cycle.g2, CD3G, LYZ, C2, `HLA-A`)) %>%
  mutate_at(c(cycle.g2, CD3G, LYZ, C2, `HLA-A`), scale) %>%

  # Plot
  ggplot(aes(UMAP_1, UMAP_2, color=value)) +
  geom_point() +
  facet_wrap(~name)

```