
**Brings Seurat to the tidyverse!**

website: [stemangiola.github.io/tidyseurat/](https://stemangiola.github.io/tidyseurat/)

Please also have a look at

- [tidyseurat](https://stemangiola.github.io/tidyseurat/) for tidy single-cell RNA sequencing analysis
- [tidySummarizedExperiment](https://tidyomics.github.io/tidySummarizedExperiment/) for tidy bulk RNA sequencing analysis
- [tidybulk](https://tidyomics.github.io/tidybulk/) for tidy bulk RNA-seq analysis
- [tidygate](https://github.com/stemangiola/tidygate/) for adding custom gate information to your tibble
- [tidyHeatmap](https://stemangiola.github.io/tidyHeatmap/) for heatmaps produced with tidy principles


```{r, echo=FALSE, include=FALSE, }
library(knitr)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

![visual cue](`r visual_cue`)

# Introduction

tidyseurat provides a bridge between the Seurat single-cell package [@butler2018integrating; @stuart2019comprehensive] and the tidyverse [@wickham2019welcome]. It creates an invisible layer that enables viewing the
Seurat object as a tidyverse tibble, and provides Seurat-compatible *dplyr*, *tidyr*, *ggplot* and *plotly* functions.


## Functions/utilities available

Seurat-compatible Functions | Description
------------ | -------------
`all` | 

tidyverse Packages | Description
------------ | -------------
`dplyr` | All `dplyr` APIs like for any tibble
`tidyr` | All `tidyr` APIs like for any tibble
`ggplot2` | `ggplot` like for any tibble
`plotly` | `plot_ly` like for any tibble

Utilities | Description
------------ | -------------
`tidy` | Add `tidyseurat` invisible layer over a Seurat object
`as_tibble` | Convert cell-wise information to a `tbl_df`
`join_features` | Add feature-wise information, returns a `tbl_df`
`aggregate_cells`| Aggregate cell gene-transcription abundance as pseudobulk tissue                  |

## Installation

From CRAN
```{r eval=FALSE}
install.packages("tidyseurat")
```

From Github (development)
```{r, eval=FALSE}
devtools::install_github("stemangiola/tidyseurat")
```

```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(magrittr)
library(ggplot2)
library(Seurat)
library(tidyseurat)
```


## Create `tidyseurat`, the best of both worlds!

This is a seurat object but it is evaluated as tibble. So it is fully compatible both with Seurat and tidyverse APIs.

```{r}
pbmc_small = SeuratObject::pbmc_small
```

**It looks like a tibble**

```{r}
pbmc_small
```

**But it is a Seurat object after all**

```{r}
pbmc_small@assays
```

# Preliminary plots

Set colours and theme for plots.

```{r}
# Use colourblind-friendly colours
friendly_cols <- c("#88CCEE", "#CC6677", "#DDCC77", "#117733", "#332288", "#AA4499", "#44AA99", "#999933", "#882255", "#661100", "#6699CC")

# Set theme
my_theme <-
  list(
    scale_fill_manual(values = friendly_cols),
    scale_color_manual(values = friendly_cols),
    theme_bw() +
      theme(
        panel.border = element_blank(),
        axis.line = element_line(),
        panel.grid.major = element_line(size = 0.2),
        panel.grid.minor = element_line(size = 0.1),
        text = element_text(size = 12),
        legend.position = "bottom",
        aspect.ratio = 1,
        strip.background = element_blank(),
        axis.title.x = element_text(margin = margin(t = 10, r = 10, b = 10, l = 10)),
        axis.title.y = element_text(margin = margin(t = 10, r = 10, b = 10, l = 10))
      )
  )
```

We can treat `pbmc_small` effectively as a normal tibble for plotting.

Here we plot number of features per cell.

```{r plot1}
pbmc_small %>%
  ggplot(aes(nFeature_RNA, fill = groups)) +
  geom_histogram() +
  my_theme
```

Here we plot total features per cell.

```{r plot2}
pbmc_small %>%
  ggplot(aes(groups, nCount_RNA, fill = groups)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.1) +
  my_theme
```

Here we plot abundance of two features for each group.

```{r}
pbmc_small %>%
  join_features(features = c("HLA-DRA", "LYZ"), shape = "long") %>%
  ggplot(aes(groups, .abundance_RNA + 1, fill = groups)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(aes(size = nCount_RNA), alpha = 0.5, width = 0.2) +
  scale_y_log10() +
  my_theme
```

# Preprocess the dataset

Also you can treat the object as Seurat object and proceed with data processing.

```{r preprocess}
pbmc_small_pca <-
  pbmc_small %>%
  SCTransform(verbose = FALSE) %>%
  FindVariableFeatures(verbose = FALSE) %>%
  RunPCA(verbose = FALSE)

pbmc_small_pca
```


If a tool is not included in the tidyseurat collection, we can use `as_tibble` to permanently convert `tidyseurat` into tibble.

```{r pc_plot}
pbmc_small_pca %>%
  as_tibble() %>%
  select(contains("PC"), everything()) %>%
  GGally::ggpairs(columns = 1:5, ggplot2::aes(colour = groups)) +
  my_theme
```

# Identify clusters

We proceed with cluster identification with Seurat.

```{r cluster}
pbmc_small_cluster <-
  pbmc_small_pca %>%
  FindNeighbors(verbose = FALSE) %>%
  FindClusters(method = "igraph", verbose = FALSE)

pbmc_small_cluster
```

Now we can interrogate the object as if it was a regular tibble data frame.

```{r cluster count}
pbmc_small_cluster %>%
  count(groups, seurat_clusters)
```

We can identify cluster markers using Seurat.

<!-- If this is Seurat v4, comment out the v3 markers -->

`r if (packageVersion("Seurat") >= package_version("4.0.0")) {"<!--"}`

```{r markers_v3, eval=(packageVersion("Seurat") < package_version("4.0.0"))}
# Identify top 10 markers per cluster
markers <-
  pbmc_small_cluster %>%
  mutate(orig.ident = seurat_clusters) %>% 
  FindAllMarkers(only.pos = TRUE) %>%
  group_by(cluster) %>%
  top_n(10, avg_logFC)

# Plot heatmap
pbmc_small_cluster %>%
  DoHeatmap(
    features = markers$gene,
    group.colors = friendly_cols
  )
```

`r if (packageVersion("Seurat") >= package_version("4.0.0")) {"-->"}`

<!-- If this is Seurat v3, comment out the v4 markers -->

`r if (packageVersion("Seurat") < package_version("4.0.0")) {"<!--"}`

```{r markers_v4, eval=FALSE}
# Identify top 10 markers per cluster
markers <-
  pbmc_small_cluster %>%
  FindAllMarkers(only.pos = TRUE, min.pct = 0.25, thresh.use = 0.25) %>%
  group_by(cluster) %>%
  top_n(10, avg_log2FC)

# Plot heatmap
pbmc_small_cluster %>%
  DoHeatmap(
    features = markers$gene,
    group.colors = friendly_cols
  )
```

`r if (packageVersion("Seurat") < package_version("4.0.0")) {"-->"}`

# Reduce dimensions

We can calculate the first 3 UMAP dimensions using the Seurat framework.

```{r umap, eval=FALSE}
pbmc_small_UMAP <-
  pbmc_small_cluster %>%
  RunUMAP(reduction = "pca", dims = 1:15, n.components = 3L)
```

And we can plot them using 3D plot using plotly.

```{r umap plot, eval=FALSE}
pbmc_small_UMAP %>%
  plot_ly(
    x = ~`UMAP_1`,
    y = ~`UMAP_2`,
    z = ~`UMAP_3`,
    color = ~seurat_clusters,
    colors = friendly_cols[1:4]
  )
```
![screenshot plotly](`r screenshot`)


## Cell type prediction

We can infer cell type identities using *SingleR* [@aran2019reference] and manipulate the output using tidyverse.

```{r eval=FALSE}
# Get cell type reference data
blueprint <- celldex::BlueprintEncodeData()

# Infer cell identities
cell_type_df <-
  GetAssayData(pbmc_small_UMAP, slot = 'counts', assay = "SCT") %>%
  log1p() %>%
  Matrix::Matrix(sparse = TRUE) %>%
  SingleR::SingleR(
    ref = blueprint,
    labels = blueprint$label.main,
    method = "single"
  ) %>%
  as.data.frame() %>%
  as_tibble(rownames = "cell") %>%
  select(cell, first.labels)
```

```{r, eval=FALSE}
# Join UMAP and cell type info
pbmc_small_cell_type <-
  pbmc_small_UMAP %>%
  left_join(cell_type_df, by = "cell")

# Reorder columns
pbmc_small_cell_type %>%
  select(cell, first.labels, everything())
```

We can easily summarise the results. For example, we can see how cell type classification overlaps with cluster classification.

```{r, eval=FALSE}
pbmc_small_cell_type %>%
  count(seurat_clusters, first.labels)
```

We can easily reshape the data for building information-rich faceted plots.

```{r eval=FALSE}
pbmc_small_cell_type %>%

  # Reshape and add classifier column
  pivot_longer(
    cols = c(seurat_clusters, first.labels),
    names_to = "classifier", values_to = "label"
  ) %>%

  # UMAP plots for cell type and cluster
  ggplot(aes(UMAP_1, UMAP_2, color = label)) +
  geom_point() +
  facet_wrap(~classifier) +
  my_theme
```

We can easily plot gene correlation per cell category, adding multi-layer annotations.

```{r eval=FALSE}
pbmc_small_cell_type %>%

  # Add some mitochondrial abundance values
  mutate(mitochondrial = rnorm(n())) %>%

  # Plot correlation
  join_features(features = c("CST3", "LYZ"), shape = "wide") %>%
  ggplot(aes(CST3 + 1, LYZ + 1, color = groups, size = mitochondrial)) +
  geom_point() +
  facet_wrap(~first.labels, scales = "free") +
  scale_x_log10() +
  scale_y_log10() +
  my_theme
```

#  Nested analyses

A powerful tool we can use with tidyseurat is `nest`. We can easily perform independent analyses on subsets of the dataset. First we classify cell types in lymphoid and myeloid; then, nest based on the new classification

```{r eval=FALSE}
pbmc_small_nested <-
  pbmc_small_cell_type %>%
  filter(first.labels != "Erythrocytes") %>%
  mutate(cell_class = if_else(`first.labels` %in% c("Macrophages", "Monocytes"), "myeloid", "lymphoid")) %>%
  nest(data = -cell_class)

pbmc_small_nested
```

Now we can independently for the lymphoid and myeloid subsets (i) find variable features, (ii) reduce dimensions, and (iii) cluster using both tidyverse and Seurat seamlessly.

```{r eval=FALSE}
pbmc_small_nested_reanalysed <-
  pbmc_small_nested %>%
  mutate(data = map(
    data, ~ .x %>%
      FindVariableFeatures(verbose = FALSE) %>%
      RunPCA(npcs = 10, verbose = FALSE) %>%
      FindNeighbors(verbose = FALSE) %>%
      FindClusters(method = "igraph", verbose = FALSE) %>%
      RunUMAP(reduction = "pca", dims = 1:10, n.components = 3L, verbose = FALSE)
  ))

pbmc_small_nested_reanalysed
```

Now we can unnest and plot the new classification.

```{r eval=FALSE}
pbmc_small_nested_reanalysed %>%

  # Convert to tibble otherwise Seurat drops reduced dimensions when unifying data sets.
  mutate(data = map(data, ~ .x %>% as_tibble())) %>%
  unnest(data) %>%

  # Define unique clusters
  unite("cluster", c(cell_class, seurat_clusters), remove = FALSE) %>%

  # Plotting
  ggplot(aes(UMAP_1, UMAP_2, color = cluster)) +
  geom_point() +
  facet_wrap(~cell_class) +
  my_theme
```

#  Aggregating cells 

Sometimes, it is necessary to aggregate the gene-transcript abundance from a group of cells into a single value. For example, when comparing groups of cells across different samples with fixed-effect models.

In tidyseurat, cell aggregation can be achieved using the `aggregate_cells` function.
 
 
```{r, eval=FALSE}
pbmc_small %>%
  aggregate_cells(groups, assays = "RNA")
```

   